Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-03-28 thesis/draft.tex
temp storage

\section{Discussion}

>>>>

\section{Discussion}
In this study, we introduce the Geometric Chain Trick (GCT), characterized by a fixed number of substages and a geometric series of transition rates. This approach aims to address the limitations identified in the Linear Chain Trick (LCT), including rigid parameter values and a cumbersome structure determination process. Our results indicate that the GCT model can utilize a uniform structure to achieve efficacy comparable to that of the LCT model with varying substage structures. The comparison is based on the consistency of simulation results, examining both the cohort model (stage duration distribution) and the transmission model (incidence over time). This enhancement simplifies the optimization of parameter values, requiring only a one-time process during data fitting. Furthermore, by theoretically analyzing and empirically validating the model through data fitting, we demonstrate that our model offers enhanced flexibility, specifically a broader range for the possible values of $\kappa$, in modeling the distribution of infectious stage duration. This flexibility allows for the selection of mean $(M)$ and shape parameters $(\kappa)$ for the distribution, with $M \in \mathbb{R}^+$ and $\kappa \in [\frac{1}{n}, 1)$.

Compared to current studies that aim to provide a more flexible dwell time distribution within the ODE framework, such as the Generalized Linear Chain Trick (GLCT) proposed by Hurtado and Kirosingh \cite{hurtado2019generalizations}, our method offers a simpler concept and structure for constructing, simulating, and analyzing models. This simplicity makes our approach more accessible to a wide range of scientists and fields of study. Additionally, our method allows for considering a more diverse substage transition pattern, for instance, a higher dimension transition matrix (e.g. Markov Chain). This idea is inspired by the GLCT \cite{hurtado2019generalizations}\cite{hurtado2021building}, yet it is less complex. In our Geometric Chain Trick (GCT), transitions between substages follow a sequential pattern, where each substage receives a unique input and progresses to another distinct substage, with no cycles among them. Modifying this pattern of transitions between substages could possibly yield a wider variety of stage duration distributions.

There are potential challenges that need to be addressed in the current GCT model. As discussed in Section 4.x, the parameters estimated after data fitting do not precisely reflect the actual values, yet these values are often crucial for understanding disease dynamics. For example, the basic reproduction number, $R_0$, which indicates how many secondary infections can be expected from a single infected individual in a fully susceptible population, is essential for assessing the potential for disease outbreak and epidemic spread. The calculation of $R_0$ requires the use of the contact rate $(\beta)$, recovery rate $(\gamma)$, and birth/death rate $(\mu)$. Therefore, achieving accurate parameter values from model fitting is necessary, and more study is needed.

In addition to seeking solutions to the challenges discussed and exploring the diverse transition patterns mentioned above, future efforts will also involve fitting the GCT model to real-world incidence and interval data to assess its performance. We also plan to integrate this framework into the SEIR (Susceptible, Exposed, Infectious, Recovered) model to better account for delayed responses. Many infectious diseases exhibit a notable incubation period during which the pathogen replicates within the host without making the host infectious. Incorporating an exposed phase into the SEIR model enables a more accurate representation of disease transmission dynamics across populations.

In conclusion, this study suggests that our method simplifies the application of a uniform model architecture for data fitting and exhibits heightened adaptability for parameter value selection within an Ordinary Differential Equation (ODE) framework. Specifically, in the field of epidemiology, our research holds the potential to improve model performance and facilitate the creation of user-friendly tools for robust epidemic forecasting and simulation. The method demonstrates considerable flexibility in its structure, making it suitable to a wide array of applications across different biological systems.
